Consider an undirected graph $G(V, E, w)$ with $V$ representing the set of vertices, $E$ the set of edges, and $w_{ij} = w_{ji}$ denoting the weight associated with each edge. In the case of an unweighted graph, we assume unit weight for each edge ($w_{ij} = 1$). Additionally, the neighbors of a vertex $i$ are denoted as $J_i = \{j\ |\ (i, j) \in E\}$, the weighted degree of each vertex as $K_i = \sum_{j \in J_i} w_{ij}$, the total number of vertices as $N = |V|$, the total number of edges as $M = |E|$, and the sum of edge weights in the undirected graph as $m = \sum_{i, j \in V} w_{ij}/2$.




\subsection{Community detection}

Disjoint community detection is the process of identifying a community membership mapping, $C: V \rightarrow \Gamma$, where each vertex $i \in V$ is assigned a community-id $c \in \Gamma$, where $\Gamma$ is the set of community-ids. We denote the vertices of a community $c \in \Gamma$ as $V_c$, and the community that a vertex $i$ belongs to as $C_i$. Further, we denote the neighbors of vertex $i$ belonging to a community $c$ as $J_{i \rightarrow c} = \{j\ |\ j \in J_i\ and\ C_j = c\}$, the sum of those edge weights as $K_{i \rightarrow c} = \sum_{j \in J_{i \rightarrow c}} w_{ij}$, the sum of weights of edges within a community $c$ as $\sigma_c = \sum_{(i, j) \in E\ and\ C_i = C_j = c} w_{ij}$, and the total edge weight of a community $c$ as $\Sigma_c = \sum_{(i, j) \in E\ and\ C_i = c} w_{ij}$ \cite{com-leskovec21}.




\subsection{Modularity}

Modularity serves as a\ignore{fitness} metric for assessing the quality of communities identified by heuristic-based community detection algorithms. It is computed as the difference between the fraction of edges within communities and the expected fraction if edges were randomly distributed, yielding a range of $[-0.5, 1]$ where higher values indicate better results \cite{com-brandes07}.\ignore{The optimization of this metric theoretically leads to the optimal grouping \cite{com-newman04, com-traag11}.} The modularity $Q$ of identified communities is determined using Equation \ref{eq:modularity}, where $\delta$ represents the Kronecker delta function ($\delta (x,y)=1$ if $x=y$, $0$ otherwise). The \textit{delta modularity} of moving a vertex $i$ from community $d$ to community $c$, denoted as $\Delta Q_{i: d \rightarrow c}$, can be computed using Equation \ref{eq:delta-modularity}.

\begin{equation}
\label{eq:modularity}
  Q
  = \frac{1}{2m} \sum_{(i, j) \in E} \left[w_{ij} - \frac{K_i K_j}{2m}\right] \delta(C_i, C_j)
  = \sum_{c \in \Gamma} \left[\frac{\sigma_c}{2m} - \left(\frac{\Sigma_c}{2m}\right)^2\right]
\end{equation}

\begin{equation}
\label{eq:delta-modularity}
  \Delta Q_{i: d \rightarrow c}
  = \frac{1}{m} (K_{i \rightarrow c} - K_{i \rightarrow d}) - \frac{K_i}{2m^2} (K_i + \Sigma_c - \Sigma_d)
\end{equation}




\subsection{Louvain algorithm}
\label{sec:about-leiden}

The Louvain method \cite{com-blondel08} is a modularity optimization based agglomerative algorithm for identifying high quality disjoint communities in large networks. It has a time complexity of $O (L |E|)$ (with $L$ being the total number of iterations performed), and a space complexity of $O(|V| + |E|)$ \cite{com-lancichinetti09}. The algorithm consists of two phases: the \textit{local-moving phase}, where each vertex $i$ greedily decides to move to the community of one of its neighbors $j \in J_i$ that gives the greatest increase in modularity $\Delta Q_{i:C_i \rightarrow C_j}$ (using Equation \ref{eq:delta-modularity}), and the \textit{aggregation phase}, where all the vertices in a community are collapsed into a single super-vertex. These two phases make up one pass, which repeats until there is no further increase in modularity \cite{com-blondel08, com-leskovec21}. We observe that Louvain obtains high-quality communities, with $3.0 - 30\%$ higher modularity than that obtained by LPA, but requires $2.3 - 14\times$ longer to converge.




\subsection{Leiden algorithm}
\label{sec:about-leiden}

%% Problems with Louvain: Internally disconnected communities
Louvain algorithm may identify communities that are internally disconnected. A certain node that acts as bridge may move to another community. Remaining vertices stay well connected. This occurs frequently in practice. Iterating the algorithm aggravates the problem. Although its does not decrease quality function.


%% Problems with Louvain: Arbitrarily badly connected communities

Further, Louvain algorithm may identify communities that are arbitrarily badly connected. Nodes that are grouped together may be better off split into multiple communities, i.e., splitting the community can improve the quality. This is not the same as resolution limit problem with modularity, that causes small communities to be clustered with large communities. Louvain only guarantees that no communities can be merged (well separated). Badly connected => Leiden could split given communities.


%% Leiden algorithm

Unlike the Louvain algorithm, the Leiden algorithm has an additional refine phase after the local-moving phase, where nodes within each community are local-moved again after reset. Refine phase is randomized (not greedy), the community a node moves to is proportional to the delta-modularity of moving to that community. When nodes are greedily moved, some optimal partitions cannot be found. It uses local-move with pruning (however, we apply this optimizations to both algorithms).


%% Leiden algorithm: Guarantees

1. All communities are $\gamma$-separated.
2. All communities are $\gamma$-connected.
If partition didn’t change:
3. All nodes are optimally assigned.
4. All communities are subpartition $\gamma$-dense.
5. All communities are uniformly $\gamma$-dense.
6. All communities are subset optimal.


%% Idea behind Leiden algorithm

The idea behind the new leiden algorithm is that louvain can find arbitrarily badly connected communities and even disconnected communities. Louvain algorithm has two phases, a locl-moving phase, and an aggregation phase (for each pass). Leiden algorithm adds another refinement phase between the two, where it allows vertices to find sub-communities within the communities just obtained from local-moving phase.


%% Came across Leiden algorithm

I came across this paper on Leiden method which shows that some communities produced by Louvain method are badly connected, and some even disconnected (and why this is not good). Louvain method relies on modularity score, which has resolution issues (cant handle too small communities, solution CPM score). They add a refinement step and a better (random) local move approach. The restructuring phase is also with refinement. They also proved is supplementary section why its better.
https://www.nature.com/articles/s41598-019-41695-z

In both:
Phase 1: Iterate until the score no longer improves (but Leiden can still proceed, maybe because its random)
Phase 2: Form its smaller super-nodes graph
Repeat until score no longer improves


%% More about Leiden algorithm

- Separating disconnected communities
Louvain algorithm may place diconnected groups of vertices within the same community. We could try disconnecting them either at the end or at the end of each phase.
- Dynamic Leiden algorithm
This involves applying a refinement phase after the local-moving phase of the Louvain algorithm. It also should be able to handle community split without a call to Static Louvain/Leiden.
In the end we can measure the quality of communities obtained through our disconnected Louvain and dynamic Leiden in terms of quality, stability, and performance.
